---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
options(readr.show_col_types = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
```

# datagouv

<!-- badges: start -->
[![R-CMD-check](https://github.com/astamm/datagouv/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/astamm/datagouv/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/astamm/datagouv/graph/badge.svg)](https://app.codecov.io/gh/astamm/datagouv)
[![pkgdown](https://github.com/astamm/datagouv/actions/workflows/pkgdown.yaml/badge.svg)](https://github.com/astamm/datagouv/actions/workflows/pkgdown.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/datagouv)](https://CRAN.R-project.org/package=datagouv)
<!-- badges: end -->

`datagouv` is an R client for the public API of
[data.gouv.fr](https://www.data.gouv.fr), the French government's open data
platform. It helps you *find* a dataset that matches your interests, *judge*
whether it is usable, *download* it, and later *re-fetch the exact same table*
reproducibly. Requests are built on top of the
[`httr2`](https://httr2.r-lib.org) package.

## Installation

```r
# From GitHub once the package is published
# remotes::install_github("astamm/datagouv")
```

## Try it in seconds

The examples below hit the live data.gouv.fr API and show real results.

Find datasets matching a topic — `dg_list_datasets()` searches titles and
descriptions server-side:

```{r}
library(datagouv)

hits <- dg_list_datasets(q = "vélo", n = 5)
hits[, c("title", "id", "organization", "quality_score", "views")]
```

`id` is the stable identifier you use to download; `quality_score` and
`views` are among the rich per-dataset metadata the v2 search API embeds
inline (see `license`, `access_type`, `frequency`, `temporal_start`/`end` and
`featured` too). Because v2 search does not inline a dataset's resources, the
resource-based columns `n_resources`, `formats`, `has_table` and `has_schema`
are `NA` by default; pass `resources = TRUE` to opt into the per-dataset
resource fetch and fill them exactly. You can restrict the catalog to datasets
that carry at least one resource in a given format — e.g. only the more compact
`parquet` files, which are quicker to download:

```{r}
parquet_only <- dg_list_datasets(format = "parquet", n = 5)
parquet_only$title
```

Glimpse a dataset's health and engagement metadata before deciding to pull it.
`dg_glimpse()` surfaces the v2-inline `quality` score and flags, `metrics`
(views, downloads, followers, ...) and context that the fetch path does not
expose:

```{r}
g <- dg_glimpse("6a6be5976a05df136d48fb7a")
g$quality
g$metrics
```

Download a dataset and inspect it. The result is a single table whose stable
address is attached as an `id` attribute:

```{r}
tbl <- dg_pull_dataset("6397c0ff56d3963118a18345")     # C.vélo bike stations
head(tbl[, 1:6])
dg_table_id(tbl)
```

Judge whether the columns mean what you think. `dg_schema()` resolves the
dataset's declared schema and returns human-readable column documentation (pass
the table directly — its id is read automatically):

```{r}
# Find a schema-documented dataset, then look at its documented columns.
irve <- dg_pull_dataset("6a84778d27ac6d44d5fabe1f")     # IRVE charging points
dg_schema(irve)[, c("name", "description", "type", "example")]
```

Re-fetch the exact same table later, reproducibly, from its stored id — no
matter how the catalog changes underneath you:

```{r}
again <- dg_refetch(tbl)

identical(again, tbl)
```

And compute summary metrics — size, columns, rows and missing-value rate — on
one table or several at once:

```{r}
dg_summarise(datasets = list(iris = iris, mtcars = mtcars))
```

## Supported formats

`dg_pull_dataset()` downloads the first tabular resource of a dataset among
CSV, CSV.GZ, XLS, XLSX, PARQUET, TSV, TXT and JSON, and returns the parsed
table as a single tibble. A ZIP resource is unpacked and its first parseable
file is returned by default; `all_files = TRUE` keeps every contained file in
one of these formats as a named list — one element per file. The
delimiter of CSV/TXT resources is auto-detected (comma, semicolon, tab, pipe,
...), so both standard and European-style (semicolon/comma-decimal) files are
handled without special configuration.

The discovery catalog (`dg_list_datasets()`) is restricted to the official
tabular formats data.gouv.fr itself indexes (`csv`, `csv.gz`, `xls`, `xlsx`,
`parquet`), so every listed dataset is in principle openable as a table. Use
the `format` argument to narrow the catalog to datasets with a resource in a
specific format (the v2 search API matches several formats as a server-side
union). The exact `n_resources`/`formats`/`has_table`/`has_schema` resource
columns are populated by passing `resources = TRUE`. Direct pulls additionally
accept `tsv`, `txt` and `json` resources.

See the [vignette](https://astamm.github.io/datagouv/articles/datagouv.html) for
the full workflow and API reference.

## Disclaimer

This package is not affiliated with or endorsed by data.gouv.fr. It is an
independent project, and the authors are not responsible for the content of
the datasets it indexes. This first version is a proof of concept and may
contain bugs. Please report any issues on the
[GitHub repository](https://github.com/astamm/datagouv/issues).

This package is the result of a joint effort initiated during the French
[Finist'R](https://stateofther.pages-forge.inrae.fr/finistr2026/) bootcamp which
was held in Roscoff in August 2026. It served as a complex task to benchmark R
package development assisted by AI. The refinements that came up during the
bootcamp were made with the assistance of `deepseek-v4-flash` as provided by
Albert API, the French government's AI platform. The author is grateful to the
Finist'R organizers and the Albert API team for their support and guidance.