---
title: "Finding, judging and re-fetching French open data with datagouv"
author: "Aymeric Stamm"
format: html
editor: source
vignette: >
  %\VignetteIndexEntry{Finding, judging and re-fetching French open data}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

> This vignette assumes you can reach the data.gouv.fr API. Code that touches
> the live API only runs when the vignette is rendered with `DATAGOUV_LIVE=1`
> set (as when building this site with pkgdown); it is skipped during `R CMD
> build`/`R CMD check` so those builds stay clean. The worked examples below
> show real output from a live render.

The examples that hit the live API are marked `#| live: true`; they only run
when the document is rendered with the `DATAGOUV_LIVE=1` environment variable
set (the pkgdown site build sets it), and are skipped otherwise — in particular
during `R CMD build`/`R CMD check`, which render this vignette in a subprocess
where the usual `_R_CHECK_PACKAGE_NAME_` marker is *not* set and therefore
cannot be relied on to suppress live code. In-memory examples run
unconditionally:

```{r}
#| include: false
has_datagouv <- requireNamespace("datagouv", quietly = TRUE)
if (has_datagouv) {
  library(datagouv)
}
library(dplyr)     # for the pipe workflow (pull(), head())
knitr::opts_hooks$set(
  live = function(options) {
    if (isTRUE(options$live)) {
      # Evaluate only when explicitly opted in. Fail closed: absent or unset
      # DATAGOUV_LIVE means "do not touch the network", which is what R CMD
      # build/check (and any ordinary render) want.
      options$eval <- toupper(Sys.getenv("DATAGOUV_LIVE")) == "1"
    }
    options
  },
  dg = function(options) {
    # Gate the in-memory, network-free example chunks (marked `#| dg: <fn>`)
    # so a broken install can never abort R CMD build/check. We fail closed:
    # the chunk runs only on the website — pkgdown sets DATAGOUV_LIVE=1, where
    # the package is guaranteed to work — and is skipped during the packaging
    # build, which normally does not set that variable. A requireNamespace()/
    # get() "is it callable?" prediction proved unreliable on Windows/R-devel,
    # where a present-but-unforceable lazy-load export reports callable yet
    # throws when the function is actually invoked, hard-failing the build
    # (see AGENTS.md). Gating on DATAGOUV_LIVE is therefore a best-effort
    # guard: even with the chunk allowed to run, it can no longer abort the
    # render, because its body is wrapped in try() (a low-level catch-all
    # that contains even the unforceable-export hard failure) and it sets
    # `#| error: true`. dg_summary()/dg_summarise() are still fully exercised
    # by the test suite.
    fun <- options$dg
    ok <- is.character(fun) && length(fun) == 1L &&
      toupper(Sys.getenv("DATAGOUV_LIVE")) == "1" &&
      "package:datagouv" %in% search() &&
      !is.null(tryCatch(get(fun, inherits = TRUE), error = function(e) NULL))
    options$eval <- ok && isTRUE(options$eval)
    options
  }
)
```

## The problem datagouv solves

`datagouv` is a small R client for the public API of
[data.gouv.fr](https://www.data.gouv.fr), the French government's open data
platform. It was written with a specific user in mind: a student or a data
scientist who wants to *find* a dataset that matches their interests, *judge*
whether it is usable for analysis, *download* it, and later *re-fetch the exact
same table* in a reproducible way.

These four steps may sound trivial, but the platform makes each of them harder
than it should be:

1. **Finding.** Enumerating "all" published datasets means paging through
   tens of thousands of records. The catalog search is server-side and only
   matches titles and descriptions, which are free-form and often uninformative.
2. **Judging.** The tabular API provides no human-readable variable
   descriptions. The actual descriptions live in separate *schema* documents on
   [schema.data.gouv.fr](https://schema.data.gouv.fr); data.gouv.fr only attaches
   a pointer (`name` / `url`) to a resource. Without resolving that pointer, you
   cannot tell whether a column really means what you assume.
3. **Downloading.** Datasets come in many formats (CSV, Excel, ZIP, Parquet,
   JSON...), the declared format is not always accurate, and a "dataset" may
   contain several files.
4. **Re-fetching.** Catalog titles and even ordering change over time, so a
   script that reaches for a dataset "by title" is fragile. You want an address
   that always resolves to the same table.

`datagouv` addresses all four. Its functions are organised around that workflow:

| Step | Function |
|------|----------|
| Find / search the catalog | `dg_find_datasets()`, `dg_find_organization()`, `dg_find_topics()` |
| Judge documented columns | `dg_schema()` |
| Download tabular resources | `dg_pull_dataset()` |
| Summarise table contents | `dg_summary()`, `dg_summarise()` |
| Re-fetch a table reproducibly | `dg_refetch()` |

The approach mirrors what several US cities propose (e.g.
[`nycOpenData`](https://github.com/ropensci/nycOpenData) for New York), but
tailored to the data.gouv.fr API.

## Finding datasets

`dg_find_datasets()` returns a tibble with one row per dataset:

```{r}
#| live: true
library(datagouv)

datasets <- dg_find_datasets(n = 20)
head(datasets)
```

The columns are chosen to help you decide, at a glance, whether a dataset is
worth pulling:

* `id` is the stable, unique identifier used to address the dataset later.
* `n_resources` is the number of files the dataset contains.
* `formats` lists the distinct file formats found among them.
* `has_table` is `TRUE` when at least one resource can be parsed into a table
  by this package.
* `has_schema` is `TRUE` when at least one resource declares a data schema,
  i.e. when per-variable documentation is available via `dg_schema()`.

The search endpoint does not inline each dataset's resources, so the
resource-based columns `n_resources`, `formats`, `has_table` and `has_schema`
are `NA` unless you opt in with `resources = TRUE` (which costs one extra
request per dataset):

```{r}
#| live: true
cycle <- dg_find_datasets(q = "vélo", n = 10, resources = TRUE)
cycle[, c("title", "n_resources", "has_table", "has_schema")]
```

The discovery catalog is **restricted to data.gouv's official tabular
formats** (`csv`, `csv.gz`, `xls`, `xlsx`, `parquet`), so every listed dataset
is in principle openable as a table — `has_table` is almost always `TRUE`. This
is a deliberate choice: JSON, TSV and TXT resources can still be parsed when
you address them directly (see `dg_pull_dataset()`), but they are not guaranteed
tabular and are left out of the catalog so that "the catalog" stays a reliable
list of tables.

### Restricting to documented datasets

Because descriptions live in schemas and only a fraction of datasets declare
one, `dg_schema()` only helps on a subset of the catalog. You can target that
subset directly. v2 has no boolean "declares a schema" server-side filter, so
`schema_only` filters client-side on `has_schema`, which needs the per-dataset
resource fetch — so calling it forces `resources = TRUE` (with a message about
the extra requests) and the filter just works:

```{r}
#| live: true
documented <- dg_find_datasets(schema_only = TRUE, n = 10)
documented[, c("title", "has_schema")]
```

### Restricting to specific formats

You can narrow the catalog to datasets that carry a resource in a format of
your choice with the `format` argument. This is especially useful to find
lighter files (e.g. `parquet`) that download faster than their CSV twins:

```{r}
#| live: true
parquet <- dg_find_datasets(format = "parquet", n = 10)
parquet[, c("title", "formats")]
```

Multiple formats can be requested at once; each is queried server-side and the
results are combined.

### Finding a specific organization's datasets

`dg_find_datasets(organization =)` can address a producer by its 24-hex id, its
`name`, or its `slug`. `dg_find_organization()` lists the organizations known to
data.gouv so you can discover which one you want and get its stable id:

```{r}
#| live: true
orgs <- dg_find_organization(q = "SNCF")
orgs[, c("name", "slug", "datasets")]
```

The slug resolves to the same datasets as the corresponding id — pass either to
`dg_find_datasets()`:

```{r}
#| live: true
sncf <- dg_find_datasets(organization = "sncf", n = 10)
sncf[, c("title", "organization")]
```

### Finding datasets grouped under a theme

Beyond producers, data.gouv curates datasets into *themes* (topics) such as
"Mobilité", "Environnement" or "Énergie". `dg_find_topics()` lists these
themes — including how many elements each groups — so you can discover one and
get its stable 24-hex id:

```{r}
#| live: true
topics <- dg_find_topics(q = "mobilité", n = 5)
topics[, c("name", "n_elements")]
```

Pass a theme's id to `dg_find_datasets(topic =)` to narrow a catalog search to
datasets grouped under it (the same single-valued server-side filter that
`organization`/`geozone` use; it takes a topic id, not a name/slug):

```{r}
#| live: true
mobility <- dg_find_datasets(topic = topics$id[1], n = 10)
mobility[, c("title", "organization")]
```

By default `dg_find_topics()` reports `n_elements` — the theme's declared total
element count (datasets, reuses, dataservices, ...). To see how that total
breaks down by kind, pass `elements = TRUE` (costing one extra request per
topic); the `n_datasets`/`n_dataservices`/`n_reuses` columns are `NA`
otherwise.

## Judging whether a dataset is usable

The judged usefulness of a dataset hinges on whether the columns mean what you
think they mean. That information comes from the producer's *schema* on
schema.data.gouv.fr. `dg_schema()` takes a table (its id is read automatically)
or a composed table id and returns the documented fields:

```{r}
#| live: true
# schema_only filters client-side, so request a batch and take the first hit.
documented <- dg_find_datasets(schema_only = TRUE, n = 100)
table_id <- documented$id[!is.na(documented$id)][[1]]

# Pull it, then inspect the schema of the returned table.
tbl <- dg_pull_dataset(table_id)
schema <- dg_schema(tbl)

# Human-readable titles and descriptions of every column:
head(schema)
```

The result is a tibble with one row per column and the columns `name`,
`title`, `description`, `type` and `example`, together with the schema's own
`title` and `name` attached as attributes. Where the schema provides no title
or description (some producers document only some fields), the corresponding
cell is `NA`.

If the resource carries no schema pointer, `dg_schema()` returns `NULL` with a
message explaining that no variable documentation is available.

## Downloading data

`dg_pull_dataset()` downloads the first parseable tabular resource of a dataset
and returns a **single tibble**:

```{r}
#| live: true
tbl <- dg_pull_dataset("6397c0ff56d3963118a18345")
head(tbl)
dg_table_id(tbl)
```

A few things to know about pulling:

* Supported formats are `csv`, `csv.gz`, `xls`, `xlsx`, `parquet`, `tsv`, `txt`
  and `json`. A **ZIP** resource is unpacked and its first parseable file is
  returned by default; `all_files = TRUE` keeps every contained file in one of
  these formats as a named list — one element per file.
* The delimiter of CSV/TXT resources is auto-detected (comma, semicolon, tab,
  pipe, ...), so both standard and European-style files (semicolon /
  decimal-comma) are handled without configuration.
* Declared formats are not always accurate, and a candidate may fail to parse
  (e.g. a `json` resource that actually serves an API metadata document).
  `dg_pull_dataset()` skips non-parseable resources and falls back to the next
  tabular one instead of erroring.
* When a dataset offers the *same table* in several formats (same file name,
  different extension, e.g. `data.csv` vs `data.xlsx`), the lightest advertised
  file is downloaded so the pull is as small as possible; resources with
  distinct names keep their declared order.
* Every returned table carries its stable, unique address as an `id`
  attribute, readable with `dg_table_id()`.

## Re-fetching the same table reproducibly

This is what makes your analysis reproducible over time. The table's id is a
URI built from the platform's own stable identifiers —
`https://www.data.gouv.fr/datasets/<dataset_id>#<resource_id>` (plus
`/<file>` for a file inside a ZIP). Unlike a human-readable title, this
address always resolves to the same table:

```{r}
#| live: true
tbl <- dg_pull_dataset("6397c0ff56d3963118a18345")
table_id <- dg_table_id(tbl)
table_id

# Re-fetch the exact same table later:
again <- dg_refetch(tbl)
```

`dg_refetch()` accepts the table id (URI) directly, so you can store it in a
script or a database and reproduce the pull without re-searching the catalog.

## Summarising datasets

`dg_summary()` computes metrics for a single table:

```{r}
#| dg: dg_summary
#| error: true
# The call is wrapped in try() as a low-level backstop: neither knitr's
# `error: true` option nor the DATAGOUV_LIVE/get() gate can contain a
# present-but-unforceable lazy-load export (the Windows/R-devel failure,
# see AGENTS.md), whereas try() degrades even that hard failure to printed
# output instead of aborting R CMD build/check.
try(dg_summary(iris, name = "iris"))
```

The reported columns are `dataset` (a label), `size_kb` (in-memory weight),
`n_vars`, `n_numeric`, `n_non_numeric`, `n_rows` and `prop_missing` (the
proportion of missing values). A table's id is carried as an attribute, not a
column, so it never inflates these metrics.

`dg_summarise()` applies `dg_summary()` to a collection of tables. It is
flexible about its input, accepting:

* a named list of tibbles (each element is a single table),
* a named list of such lists, as returned by `dg_pull_dataset(all_files = TRUE)`
  (a ZIP may contribute several tables),
* a tibble from `dg_find_datasets()` (each dataset is downloaded and
  summarised),
* a character vector of identifiers (or exact titles),
* or `NULL` to download and summarise the first `n` datasets of the catalog.

```{r}
# In-memory tables — no network needed
#| dg: dg_summarise
#| error: true
try(dg_summarise(datasets = list(iris = iris, mtcars = mtcars)))
```

## A complete workflow

Because every step returns something the next one can consume, the whole
"find → judge → fetch" pipeline can be written as a single pipe. See how the
table flows from one step to the next without any intermediate variables:

```{r}
#| live: true
# Find a dataset, take its first id, pull it into a table and read its schema.
dg_find_datasets(q = "recharge électrique", schema_only = TRUE, n = 5) |>
  pull(id) |>
  head(1) |>
  dg_pull_dataset() |>
  dg_schema()
```

`dg_pull_dataset()` always returns a single tibble (a ZIP yields its first
parseable file), so the pipe keeps flowing whether or not the dataset is an
archive — `dg_schema()` and `dg_refetch()` read the table's stable id from its
attribute automatically. The same id lets you reproduce the exact table later
in a fresh session, no matter how the catalog changes in the meantime:

```{r}
#| live: true
tbl <- dg_find_datasets(q = "recharge électrique", schema_only = TRUE, n = 5) |>
  pull(id) |>
  head(1) |>
  dg_pull_dataset()

# Save the stable address, then re-fetch the exact same table later.
tbl_id <- dg_table_id(tbl)
again <- dg_refetch(tbl_id)
identical(again, tbl)
```

The table id is the key to reproducibility: save `tbl_id`, and
`dg_refetch(tbl_id)` returns the same table again, regardless of filename
reorganisation or later edits to the dataset on the platform.
